Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-06 aistats.tex

The statistics and optimization perspective are so tightly linked in this problem, that we believe progress from one field would help the other one. 
Progress from statistics could bring forward research in non-euclidean optimization, for instance helping in the development of accelerated methods.
On the other hand, optimization can become  a more versatile and general tool for statisticians. % handwavy and shady af.

@@

Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.

@@

This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.
\RLP{Add numerical experiments with upper bound for gaussian variance MAP.}

@@

\section{"Burn In" Analysis}
Self-concordant line of work, and \citet{kakade2010learning}.

@@

There is a long line work trying the convergence rate of stochastic gradient descent.
Beyond the results of RM51, 
we now have proofs n smooth, strongly convex problems and bounded variance 
(as well as other assumptions on the noise, reviewed later)
a decreasing step-size and an averaging scheme gets 
a $1/t$ convergence rate, 
which matches the asymptotic rate of unbiased estimation,
through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler} \footnote{
Those papers assume the stochastic gradients are bounded, 
but small modifications also work for bounded variance instead.
}

@@

However, for some problems, combining decreasing step-sizes and averaging is not necessary,
even when the function is not strongly convex.
This is the case for maximum likelihood estimation and matches the asymptotic rates, 
but holds more generally, for example for linear and logistic regressions \citep{bach2013non}.

There has been recent advances in optimization that go beyond the Euclidean case, 
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
\citep{bauschke2017descent,lu2018relatively},
where instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
we use a reference function instead \fdk{maybe add a line}
and apply mirror descent \citep{nemirovski1983problem,beck2003mirror}
(also called NoLips \citep{bauschke2017descent}, something else?).

We have convergence results for the 
application of stochastic mirror descent, 
but they typically rely on assumptions that are too strong. 
\begin{itemize}
	\item The reference function is strongly convex [typical setup in online learning]
	\item The variance is uniformly bounded
\end{itemize}
Some advances have been tried
\begin{itemize}
	\item The reference function is ``close to'' a norm
	and looks like a \citet{hanzely2018fastest}
	\item 
	Attempt to adapt the ``bounded gradient at the minimum''
	\citep{dragomir2021fast}
\end{itemize}

Other to cite;
\begin{itemize}
	\item \citet{marteauferey2019}
\end{itemize}

@@

ff
\fdk{Application 
The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
For example, we haven't yet figured out the analog of Nesterov-type acceleration 
on relatively smooth and strongly convex problems, 
to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
or just from $1/T$ to $1/T^2$ in the (non-strongly) convex case.
}

\citet{dragomir2021optimal}  shows that naïve application of Bregman updates can not achieve acceleration.
The tools developed to make progress on one problem might help make progress on the other.

@@

\section{Group Meeting}
\begin{enumerate}
	\item new plots : mirror map, and gaussian trajectories
	\item the MLE is biased according to our definition. See plots. Does this definition of bias make sense ? An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean. 
	\item \citet{kakade2010learning}
	\begin{enumerate}
		\item assumption in the spirit of infinite order self-concordance only at the optimum
		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
		\item conclusion : it is worth writing down until the end what we are obtaining with self-concordance. \RLP{TODO}
		\item idea 1 : try to express some of their results with the mean parameter instead.
		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD. (exploiting the link MAP -> optimization).
	\end{enumerate}
\end{enumerate}

\section{Plan}
\begin{enumerate}
	\item intro to exponential family, and density estimation
	\item the thing we want to bound
	\item Examples : gaussian mean and gaussian variance (+other examples, just mentioned)
	\item Insight : Strongly convex case. (+ self-concordance that is not verified either)
	\item insight : bias-variance decomposition
	\item Optimization perspective : SBPP or SBG. But no analysis hold, revealing a flaw of all these techniques.
	\item Discussion : we believe finding a convergence rate would bring new tools useful to deal with common objects such as barrier losses.
\end{enumerate}

Open questions
\begin{itemize}
	\item does a base measure change anything ?
	\item is there multiple conjugate priors ?
	\item misspecified case. Do we have a   formula ? Partial answer : a sufficient condition is to have $\mu^*$ be accessible to our model, even if the true distribution is not.
\end{itemize}

>>>>

The statistics and optimization perspective are so tightly linked in this problem, that we believe progress from one field would help the other one. 
Progress from statistics could bring forward research in non-euclidean optimization, or the optimization of barrier losses, for instance helping in the development of accelerated methods.
On the other hand, optimization can become  a more versatile and general tool for statisticians. % handwavy and shady af.

@@

Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.
A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.

@@

This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.

@@

\section{"Burn In" Statistical Analysis}
Self-concordant line of work : 
\citet{kakade2010learning},
\cite{anastasiou2017bounds},
\citet{ostrovskii2021finite}.

@@

There is a long line work bounding the convergence rate of stochastic gradient descent.
Beyond the results of \citet{robbins1951stochastic}, 
we now have proofs on smooth, strongly convex problems and bounded variance 
(as well as other assumptions on the noise, reviewed later)
a decreasing step-size and an averaging scheme gets 
a $1/t$ convergence rate, 
which matches the asymptotic rate of unbiased estimation,
through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler} \footnote{
Those papers assume the stochastic gradients are bounded, 
but small modifications also work for bounded variance instead.
}

@@

However, for some problems, combining decreasing step-sizes and averaging is not necessary,
even when the function is not strongly convex.
This is the case for maximum likelihood estimation and matches the asymptotic rates, 
but holds more generally, for example for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.

There has been recent advances in optimization that go beyond the Euclidean case, 
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
\citep{birnbaum2011distributed,bauschke2017descent,lu2018relatively}.
Instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
we use a reference function instead \fdk{maybe add a line}
and apply mirror descent \citep{nemirovski1983problem,beck2003mirror}
(also called NoLips \citep{bauschke2017descent}, something else?).

We have convergence results for the 
application of stochastic mirror descent, 
but they typically rely on assumptions that are too strong. 
\begin{itemize}
	\item The reference function is strongly convex [typical setup in online learning]
	\item The variance is uniformly bounded
\end{itemize}
Some advances have been tried
\begin{itemize}
	\item The reference function is ``close to'' a norm
	and looks like a \citet{hanzely2018fastest}
	\item Attempt to adapt the ``bounded gradient at the minimum''
	\citep{dragomir2021fast}
	\item Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}.
\end{itemize}

Other work 
\citet{marteauferey2019beyond}.

@@

\fdk{Application 
The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
For example, we haven't yet figured out the analog of Nesterov-type acceleration 
on relatively smooth and strongly convex problems, 
to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
or just from $1/T$ to $1/T^2$ in the (non-strongly) convex case.
\citet{dragomir2021optimal}  shows that naïve application of Bregman updates can not achieve acceleration.
The tools developed to make progress on one problem might help make progress on the other. 
}

@@

\section{Group Meeting}
\begin{enumerate}
	\item new plots : mirror map, and gaussian trajectories
	\item the MLE is biased according to our definition. See plots. Does this definition of bias make sense ? An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean. 
	\item \citet{kakade2010learning}
	\begin{enumerate}
		\item assumption in the spirit of infinite order self-concordance only at the optimum
		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
		\item \TODO{it is worth writing down until the end what we are obtaining with self-concordance.}
		\item idea 1 : try to express some of their results with the mean parameter instead.
		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD. (exploiting the link MAP -> optimization).
	\end{enumerate}
	\item positioning between optimization and statistics.
\end{enumerate}

### 2021-10-06 main.tex
Update figures

@@

In our case, the bias variance decomposition tells us that we are interested in the covariance between primal and dual parameters, rather than the covariance of only of these -- $\Cov(\MAPt;\MAPm)$ rather that $\Var(\MAPm)$ or $\Var(\MAPt)$.

>>>>

\begin{figure}[ht]
	\centering
	\includegraphics[width=\textwidth]{{figs/gaussians/new_linear_n0=0}.pdf}
	\includegraphics[width=\textwidth]{{figs/2d/gaussian_convergence_n0=0}.pdf}
	\caption{
	The MLE is biased in primal space.
	}
	\label{fig:MLE_bias}
\end{figure}

@@

In our case, the bias variance decomposition tells us that we are interested in the covariance between primal and dual parameters, rather than any of their individual covariance -- $\Cov(\MAPt;\MAPm)$ rather that $\Var(\MAPm)$ or $\Var(\MAPt)$.
